To be FAIR: Theory Specification Needs an Update

### 2025-06-06 manuscript.Rmd
add citation
add author role

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Maximilian S. Ernst"
    affiliation: "2,4"
    role:
      - Writing – review & editing
      - Investigation
  - name: "Noah N.N. van Dongen"
    affiliation: "5"
    role:
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "6"
    role:
      - Conceptualization
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,7"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University, dept. Methodology & Statistics"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "5"
    institution   : "University of Amsterdam, the Netherlands"
  - id            : "6"
    institution   : "Ludwig-Maximilians-Universität München, Germany"
  - id            : "7"
    institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
authornote: |
  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.

@@

With regard to the state of Interoperability in psychology, 
Kurt Lewin's [-@lewinPsychologyProcessGroup1943] adage "there's nothing as practical as a good theory" paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
But, as we argued, this is not the case.
The examples of X-Interoperability offered in Table \@ref(tab:tabmeehl) illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-Interoperable within software used for analysis.
For example, Interoperable theory could be used
to select control variables for causal inference [@cinelliCrashCourseGood2022],
or to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021] which could be automatically evaluated through integration testing [@vanlissaUsingEndpointsCheck2023].
Furthermore, theories can be X-Interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
For example, it should be possible to embed a theory about emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].

An important aspect of Interoperability lies in linking FAIR theories with existing open science methodologies---most notably, preregistration.
As noted, preregistration alone has not resolved the replication crisis in the social sciences.
This raises a fundamental question: What additional value does a FAIR theory provide, and how does it complement preregistration?

One distinction is quantitative:
FAIR theories operate at a higher level of abstraction.
They articulate general principles and expectations that cut across studies, whereas preregistrations focus on specific applications, analytic strategies, and study designs.
More fundamentally, however, there is a qualitative difference.
A FAIR theory is designed to make predictions and be revised in light of empirical results; preregistration, by contrast, prescribes how a particular study will be conducted, regardless of what the outcomes turn out to be.

This difference can be illustrated with a Bayesian analogy.
A preregistration is akin to a statistical model---it formalizes assumptions about data generation but is in itself indifferent to the data.
Combining a theory with a specific sampling strategy, study design, measurement, etc. allows one to make predictions even before observing data.
However, after observing data, the theory might require revision.
In this sense, a theory is first a prior, and then after conducting a specific investigation as prescribed by the preregistration, is refined to become a posterior.
In short, a preregistration rooted in a FAIR theory not only supports transparency, but also strengthens the epistemic cycle of prediction, testing, and revision---moving us closer to a cumulative science.

>>>>

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Conceptualization
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Maximilian S. Ernst"
    affiliation: "2,4"
    role:
      - Writing – review & editing
      - Investigation
  - name: "Noah N.N. van Dongen"
    affiliation: "5"
    role:
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "6"
    role:
      - Conceptualization
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,7"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University, dept. Methodology & Statistics"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "5"
    institution   : "University of Amsterdam, the Netherlands"
  - id            : "6"
    institution   : "Ludwig-Maximilians-Universität München, Germany"
  - id            : "7"
    institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
authornote: |
  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.

@@

With regard to the state of Interoperability in psychology, 
Kurt Lewin's [-@lewinPsychologyProcessGroup1943] adage "there's nothing as practical as a good theory" paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
But, as we argued, this is not the case.
The examples of X-Interoperability offered in Table \@ref(tab:tabmeehl) illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-Interoperable within software used for analysis.
For example, Interoperable theory could be used
to select control variables for causal inference [@cinelliCrashCourseGood2022],
or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data [@peikertReproducibleResearchTutorial2021],
or to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021] which could be automatically evaluated through integration testing [@vanlissaUsingEndpointsCheck2023].
Furthermore, theories can be X-Interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
For example, it should be possible to embed a theory about emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].
